The Key to a Performant Data Parallel Compiler

### 2016-11-15 2017-spring-pldi/hsu-2017-key_data_parallel_compiler.tex
Conversion of existing work into the LaTeX format

\begin{verbatim}
      +⍀Fd∘.=⍳1+⌈⌿Fd
1 0 0 0 0
1 1 0 0 0
1 1 1 0 0
1 1 1 1 0
1 1 1 1 1
1 1 1 1 2
1 1 1 1 3
1 1 2 1 3
1 1 2 2 3
      +⍀Ed∘.=⍳1+⌈⌿Ed
1 0 0
1 1 0
1 1 1
1 1 2
1 1 3
1 2 3
1 3 3
1 3 4
1 3 5
1 3 6
\end{verbatim}

@@

\begin{verbatim}
      3 3⍴⍳9
0 1 2
3 4 5
6 7 8
      (⍳3)+⍤¯1⊢3 3⍴⍳9
0 1  2
4 5  6
8 9 10
\end{verbatim}

@@

\begin{verbatim}
      ⊢Fc←(1+Fd)↑⍤¯1⊢+⍀Fd∘.=⍳1+⌈⌿Fd
1 0 0 0 0
1 1 0 0 0
1 1 1 0 0
1 1 1 1 0
1 1 1 1 1
1 1 1 1 2
1 1 1 1 3
1 1 2 0 0
1 1 2 2 0
      ⊢Ec←(1+Ed)↑⍤¯1⊢+⍀Ed∘.=⍳1+⌈⌿Ed
1 0 0
1 1 0
1 1 1
1 1 2
1 1 3
1 2 0
1 3 0
1 3 4
1 3 5
1 3 6
\end{verbatim}

@@

\noindent
The above determines whether \verb;P; is an ancestor to \verb;C;.
The logical functions \verb;=;, \verb;∧;, and \verb;∨; are all extended 
point-wise over arrays.
The function \verb;⊢; returns its right argument.
The function \verb;(=∨0=⊢); reads as, ``equal or a zero right argument.'' 
We take this point-wise operation and reduce it with \verb;∧⌿; (For All).
This pattern is a special case of inner product, written \verb;f.g;. 
For example, \verb;+.×; is matrix multiplication. We transform our reduction and 
Boolean function to a single predicate with the inner product operator:
\begin{verbatim}
      C∧.(=∨0=⊢)P
1
\end{verbatim}

@@

\begin{verbatim}
      ⊢Fp←('F'=Ft)⌿Fc
1 0 0 0 0
1 1 1 0 0
      ⊢Ep←('E'=Et)⌿Ec
1 0 0
1 1 0
1 3 0
\end{verbatim}

@@

\begin{verbatim}
      ⊢Fcp←(Fd↑⍤¯1⊢Fc),0
0 0 0 0 0
1 0 0 0 0
1 1 0 0 0
1 1 1 0 0
1 1 1 1 0
1 1 1 1 0
1 1 1 1 0
1 1 0 0 0
1 1 2 0 0
      ⊢Ecp←(Ed↑⍤¯1⊢Ec),0
0 0 0
1 0 0
1 1 0
1 1 0
1 1 0
1 0 0
1 0 0
1 3 0
1 3 0
1 3 0
\end{verbatim}

@@

\begin{verbatim}
      Fcp∧.(=∨0=⊢)⍉Fp
0 0
1 0
1 0
1 1
1 1
1 1
1 1
1 0
1 0
      Ecp∧.(=∨0=⊢)⍉Ep
0 0 0
1 0 0
1 1 0
1 1 0
1 1 0
1 0 0
1 0 0
1 0 1
1 0 1
1 0 1
\end{verbatim}

@@

\begin{verbatim}
      (⍳3)×⍤1⊢Ecp∧.(=∨0=⊢)⍉Ep
0 0 0
0 0 0
0 1 0
0 1 0
0 1 0
0 0 0
0 0 0
0 0 2
0 0 2
0 0 2
      ⊢Ei←⌈/(⍳3)×⍤1⊢Ecp∧.(=∨0=⊢)⍉Ep
0 0 1 1 1 0 0 2 2 2
\end{verbatim}

@@

\begin{verbatim}
      ⊢Ek←Ei⌷⍤0 2⊢Ep 
1 0 0
1 0 0
1 1 0
1 1 0
1 1 0
1 0 0
1 0 0
1 3 0
1 3 0
1 3 0
\end{verbatim}

\noindent
At this point we have two values, \verb;Ek; and \verb;Fk;, which indicate the
closest containing node that we care about for each node in the tree,
using its node coordinate.

\begin{verbatim}
      Fi←⌈/(⍳2)×⍤1⊢Fcp∧.(=∨0=⊢)⍉Fp
      ⊢Fk←Fi⌷⍤0 2⊢Fp
1 0 0 0 0
1 0 0 0 0
1 0 0 0 0
1 1 1 0 0
1 1 1 0 0
1 1 1 0 0
1 1 1 0 0
1 0 0 0 0
1 0 0 0 0
\end{verbatim}

\noindent
We use these keys to compute over the AST, particularly for function lifting 
and expression flattening as demonstrated in the next section. The repeating 
pattern is to leverage node coordinates and array operations to do stackless 
reasoning about inter-node relationships. This pattern occurs often in our 
compiler and is particularly useful to enable straightforward data-parallel 
transformations over an AST.

@@

@@

If we use the Key operator with \verb;Fk;, we get the following (the monadic
use of \verb;↓; converts \verb;Fc; from a matrix to a vector of vectors):

\begin{verbatim}
      (1↓Fk){⍺ ⍵}⌸1↓Fd,Ft,Fv,⍪↓Fc
┌─────────┬─────────────────┐
│1 0 0 0 0│┌─┬─┬─┬─────────┐│
│         ││1│E│0│1 1 0 0 0││
│         │├─┼─┼─┼─────────┤│
│         ││2│F│0│1 1 1 0 0││
│         │├─┼─┼─┼─────────┤│
│         ││2│A│0│1 1 2 0 0││
│         │├─┼─┼─┼─────────┤│
│         ││3│N│7│1 1 2 2 0││
│         │└─┴─┴─┴─────────┘│
├─────────┼─────────────────┤
│1 1 1 0 0│┌─┬─┬─┬─────────┐│
│         ││3│E│0│1 1 1 1 0││
│         │├─┼─┼─┼─────────┤│
│         ││4│V│⍵│1 1 1 1 1││
│         │├─┼─┼─┼─────────┤│
│         ││4│P│+│1 1 1 1 2││
│         │├─┼─┼─┼─────────┤│
│         ││4│V│⍵│1 1 1 1 3││
│         │└─┴─┴─┴─────────┘│
└─────────┴─────────────────┘
\end{verbatim}

@@

@@

\begin{verbatim}
      (1↓Ek){⍺ ⍵}⌸1↓Ed,Et,Ev,⍪↓Ec
┌─────┬─────────────┐
│1 0 0│┌─┬─┬─┬─────┐│
│     ││1│E│0│1 1 0││
│     │├─┼─┼─┼─────┤│
│     ││1│P│÷│1 2 0││
│     │├─┼─┼─┼─────┤│
│     ││1│E│0│1 3 0││
│     │└─┴─┴─┴─────┘│
├─────┼─────────────┤
│1 1 0│┌─┬─┬─┬─────┐│
│     ││2│V│a│1 1 1││
│     │├─┼─┼─┼─────┤│
│     ││2│P│+│1 1 2││
│     │├─┼─┼─┼─────┤│
│     ││2│V│b│1 1 3││
│     │└─┴─┴─┴─────┘│
├─────┼─────────────┤
│1 3 0│┌─┬─┬─┬─────┐│
│     ││2│V│c│1 3 4││
│     │├─┼─┼─┼─────┤│
│     ││2│P│×│1 3 5││
│     │├─┼─┼─┼─────┤│
│     ││2│V│d│1 3 6││
│     │└─┴─┴─┴─────┘│
└─────┴─────────────┘
\end{verbatim}

@@

>>>>

\begin{verbatim}
    +⍀Fd∘.=⍳1+⌈⌿Fd  │     +⍀Ed∘.=⍳1+⌈⌿Ed
1 0 0 0 0           │ 1 0 0
1 1 0 0 0           │ 1 1 0
1 1 1 0 0           │ 1 1 1
1 1 1 1 0           │ 1 1 2
1 1 1 1 1           │ 1 1 3
1 1 1 1 2           │ 1 2 3
1 1 1 1 3           │ 1 3 3
1 1 2 1 3           │ 1 3 4
1 1 2 2 3           │ 1 3 5
                    │ 1 3 6
\end{verbatim}

@@

\begin{verbatim}
      3 3⍴⍳9 │       (⍳3)+⍤¯1⊢3 3⍴⍳9
0 1 2        │ 0 1  2
3 4 5        │ 4 5  6
6 7 8        │ 8 9 10
\end{verbatim}

@@

\begin{verbatim}
      Fc←(1+Fd)↑⍤¯1⊢+⍀Fd∘.=⍳1+⌈⌿Fd
      Ec←(1+Ed)↑⍤¯1⊢+⍀Ed∘.=⍳1+⌈⌿Ed
      Fc   │        Ec
1 0 0 0 0  │  1 0 0
1 1 0 0 0  │  1 1 0
1 1 1 0 0  │  1 1 1
1 1 1 1 0  │  1 1 2
1 1 1 1 1  │  1 1 3
1 1 1 1 2  │  1 2 0
1 1 1 1 3  │  1 3 0
1 1 2 0 0  │  1 3 4
1 1 2 2 0  │  1 3 5
           │  1 3 6
\end{verbatim}

@@

\noindent The above determines whether \verb;P; is an ancestor to \verb;C;. The logical functions \verb;=;, \verb;∧;, and \verb;∨; are all extended point-wise over arrays. The function \verb;⊢; returns its right argument. The function \verb;(=∨0=⊢); reads as, ``equal or a zero right argument.'' We take this point-wise operation and reduce it with \verb;∧⌿; (For All). This pattern is a special case of inner product, written \verb;f.g;. For example, \verb;+.×; is matrix multiplication. We transform our reduction and Boolean function to a single predicate with the inner product operator:

\begin{verbatim}
      C∧.(=∨0=⊢)P
1
\end{verbatim}

@@

\begin{verbatim}
      Fp←('F'=Ft)⌿Fc
      Ep←('E'=Et)⌿Ec
      Fp   │        Ep
1 0 0 0 0  │  1 0 0
1 1 1 0 0  │  1 1 0
           │  1 3 0
\end{verbatim}

@@

\begin{verbatim}
      Fcp←(Fd↑⍤¯1⊢Fc),0
      Ecp←(Ed↑⍤¯1⊢Ec),0
      Fcp  │        Ecp
0 0 0 0 0  │  0 0 0
1 0 0 0 0  │  1 0 0
1 1 0 0 0  │  1 1 0
1 1 1 0 0  │  1 1 0
1 1 1 1 0  │  1 1 0
1 1 1 1 0  │  1 0 0
1 1 1 1 0  │  1 0 0
1 1 0 0 0  │  1 3 0
1 1 2 0 0  │  1 3 0
           │  1 3 0
\end{verbatim}

@@

\begin{verbatim}
   Fcp∧.(=∨0=⊢)⍉Fp │    Ecp∧.(=∨0=⊢)⍉Ep
0 0                │ 0 0 0
1 0                │ 1 0 0
1 0                │ 1 1 0
1 1                │ 1 1 0
1 1                │ 1 1 0
1 1                │ 1 0 0
1 1                │ 1 0 0
1 0                │ 1 0 1
1 0                │ 1 0 1
                   │ 1 0 1
\end{verbatim}

@@

\begin{verbatim}
      (⍳3)×⍤1⊢Ecp∧.(=∨0=⊢)⍉Ep
0 0 0
0 0 0
0 1 0
0 1 0
0 1 0
0 0 0
0 0 0
0 0 2
0 0 2
0 0 2
      ⊢Ei←⌈/(⍳3)×⍤1⊢Ecp∧.(=∨0=⊢)⍉Ep
0 0 1 1 1 0 0 2 2 2
      ⊢Fi←⌈/(⍳2)×⍤1⊢Fcp∧.(=∨0=⊢)⍉Fp
0 0 0 1 1 1 1 0 0
\end{verbatim}

@@

\begin{verbatim}
      Fk←Fi⌷⍤0 2⊢Fp
      Ek←Ei⌷⍤0 2⊢Ep 
      Fk   │        Ek
1 0 0 0 0  │  1 0 0
1 0 0 0 0  │  1 0 0
1 0 0 0 0  │  1 1 0
1 1 1 0 0  │  1 1 0
1 1 1 0 0  │  1 1 0
1 1 1 0 0  │  1 0 0
1 1 1 0 0  │  1 0 0
1 0 0 0 0  │  1 3 0
1 0 0 0 0  │  1 3 0
           │  1 3 0
\end{verbatim}

\noindent At this point we have two values, \verb;Ek; and \verb;Fk;, which indicate the closest containing node that we care about for each node in the tree, using its node coordinate. We use these keys to compute over the AST, particularly for function lifting and expression flattening as demonstrated in the next section. The repeating pattern is to leverage node coordinates and array operations to do stackless reasoning about inter-node relationships. This pattern occurs often in our compiler and is particularly useful to enable straightforward data-parallel transformations over an AST.

@@

Each of the following passes operates over an AST that we represent as the catenation of the field vectors above including the Node Coordinates.

\begin{verbatim}
Fast←Fd,Ft,Fv,⍪↓Fc
East←Ed,Et,Ev,⍪↓Ec
\end{verbatim}

\noindent The function \verb;↓; converts the coordinate matrices from matrices to vectors of vectors, so that the coordinates can be catenated to the other fields, leading to a 4-column matrix for the ASTs, whose fourth column is a vector of vectors.

@@

Function lifting takes an AST with nested functions (func-tions appearing inside of other functions) and lifts these functions out of their local scopes and puts them all at the top level. Normally, function lifting must consider both the core operation of lifting function as well as the lexical scop-ing of the functions and variables that appear inside of a function. As seen below, our treatment uses a slightly dif-ferent method.

If we use the Key operator (\verb;⌸;) with the pair function (\verb;{⍺ ⍵};) using \verb;Fk; as the keys and \verb;Fast; as the AST, we get the following:

\begin{verbatim}
      (1↓Fk){⍺ ⍵}⌸1↓Fast
┌─────────┬─────────────────┐
│1 0 0 0 0│┌─┬─┬─┬─────────┐│
│         ││1│E│0│1 1 0 0 0││
│         │├─┼─┼─┼─────────┤│
│         ││2│F│0│1 1 1 0 0││
│         │├─┼─┼─┼─────────┤│
│         ││2│A│0│1 1 2 0 0││
│         │├─┼─┼─┼─────────┤│
│         ││3│N│7│1 1 2 2 0││
│         │└─┴─┴─┴─────────┘│
├─────────┼─────────────────┤
│1 1 1 0 0│┌─┬─┬─┬─────────┐│
│         ││3│E│0│1 1 1 1 0││
│         │├─┼─┼─┼─────────┤│
│         ││4│V│⍵│1 1 1 1 1││
│         │├─┼─┼─┼─────────┤│
│         ││4│P│+│1 1 1 1 2││
│         │├─┼─┼─┼─────────┤│
│         ││4│V│⍵│1 1 1 1 3││
│         │└─┴─┴─┴─────────┘│
└─────────┴─────────────────┘
\end{verbatim}

@@

\begin{verbatim}
      (1↓Fk){⊂(0 'F' 0 ⍺)⍪⍵}⌸1↓Fast
┌─────────────────┬─────────────────┐
│┌─┬─┬─┬─────────┐│┌─┬─┬─┬─────────┐│
││0│F│0│1 0 0 0 0│││0│F│0│1 1 1 0 0││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││1│E│0│1 1 0 0 0│││3│E│0│1 1 1 1 0││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││2│F│0│1 1 1 0 0│││4│V│⍵│1 1 1 1 1││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││2│A│0│1 1 2 0 0│││4│P│+│1 1 1 1 2││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││3│N│7│1 1 2 2 0│││4│V│⍵│1 1 1 1 3││
│└─┴─┴─┴─────────┘│└─┴─┴─┴─────────┘│
└─────────────────┴─────────────────┘ 
      hd←0 'F' 0,∘⊂⊣      ⍝ Func. Header
      dp←(1+⊢-⊃)0⌷⍤1⊢     ⍝ New Depths
      rf←'V'@('F'∘=)1⌷⍤1⊢ ⍝ F → V
      (1↓Fk)(⊂hd⍪dp,rf,2↓⍤1⊢)⌸1↓Fast
┌─────────────────┬─────────────────┐
│┌─┬─┬─┬─────────┐│┌─┬─┬─┬─────────┐│
││0│F│0│1 0 0 0 0│││0│F│0│1 1 1 0 0││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││1│E│0│1 1 0 0 0│││1│E│0│1 1 1 1 0││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││2│V│0│1 1 1 0 0│││2│V│⍵│1 1 1 1 1││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││2│A│0│1 1 2 0 0│││2│P│+│1 1 1 1 2││
│├─┼─┼─┼─────────┤│├─┼─┼─┼─────────┤│
││3│N│7│1 1 2 2 0│││2│V│⍵│1 1 1 1 3││
│└─┴─┴─┴─────────┘│└─┴─┴─┴─────────┘│
└─────────────────┴─────────────────┘
\end{verbatim}

@@

\begin{verbatim}
      (1↓Ek){⍺ ⍵}⌸1↓East
┌─────┬─────────────┐
│1 0 0│┌─┬─┬─┬─────┐│
│     ││1│E│0│1 1 0││
│     │├─┼─┼─┼─────┤│
│     ││1│P│÷│1 2 0││
│     │├─┼─┼─┼─────┤│
│     ││1│E│0│1 3 0││
│     │└─┴─┴─┴─────┘│
├─────┼─────────────┤
│1 1 0│┌─┬─┬─┬─────┐│
│     ││2│V│a│1 1 1││
│     │├─┼─┼─┼─────┤│
│     ││2│P│+│1 1 2││
│     │├─┼─┼─┼─────┤│
│     ││2│V│b│1 1 3││
│     │└─┴─┴─┴─────┘│
├─────┼─────────────┤
│1 3 0│┌─┬─┬─┬─────┐│
│     ││2│V│c│1 3 4││
│     │├─┼─┼─┼─────┤│
│     ││2│P│×│1 3 5││
│     │├─┼─┼─┼─────┤│
│     ││2│V│d│1 3 6││
│     │└─┴─┴─┴─────┘│
└─────┴─────────────┘
      hd←0'E'0,∘⊂⊣
      rf←'V'@('E'∘=)1⌷⍤1⊢
      ⊖(1↓Ek)(hd⍪dp,rf,2↓⍤1⊢)⌸1↓East
┌─┬─┬─┬─────┐
│0│E│0│1 3 0│
├─┼─┼─┼─────┤
│1│V│c│1 3 4│
├─┼─┼─┼─────┤
│1│P│×│1 3 5│
├─┼─┼─┼─────┤
│1│V│d│1 3 6│
└─┴─┴─┴─────┘
┌─┬─┬─┬─────┐
│0│E│0│1 1 0│
├─┼─┼─┼─────┤
│1│V│a│1 1 1│
├─┼─┼─┼─────┤
│1│P│+│1 1 2│
├─┼─┼─┼─────┤
│1│V│b│1 1 3│
└─┴─┴─┴─────┘
┌─┬─┬─┬─────┐
│0│E│0│1 0 0│
├─┼─┼─┼─────┤
│1│V│0│1 1 0│
├─┼─┼─┼─────┤
│1│P│÷│1 2 0│
├─┼─┼─┼─────┤
│1│V│0│1 3 0│
└─┴─┴─┴─────┘
\end{verbatim}

@@

\section{Performance}

\section{Discussion}
